Unrecognized States: Theory and Cases

### 2017-07-01 theory.tex
first pass at cutting down model section

The game begins at a status quo in which the secessionist elite controls at least some of the disputed territory, but cannot gain international recognition unless the central government cedes its claim to the territory. This condition is archetypical of cases in which a militarily successful war of secession ends in a ceasefire.

@@

\item $c$ chooses an investment level $R_c^t\in \mathbb{R}^+$ to influence the payoffs of $s$ and $g$.\footnote{We assume that player $c$, which is often a large coalition of states, does not have a binding budget constraint.}

@@

The game continues into the next period if the `inside players'---the secessionists and government---selected (status quo, status quo) or (cede, cede). Otherwise, the game ends in an absorbing state as described in Section~\ref{sec:stage}. Both because of the presence of absorbing states and state variables, this is a dynamic game rather than a standard repeated game.

Future payoffs are discounted with a common parameter $\de$, where 1/$\de$ is each player's discount rate and $0\leq\de\leq1$.\footnote{All of the results that follow go through essentially unchanged in the general case where the players potentially discount at different rates. This simplifying assumption is made solely to highlight the essential mechanisms of the model.} Therefore payoffs for the entire game for player $i\in \{s, g, p, c\} $ can be expressed by the discounted stream of payments $\Sigma_{t=1}^{\infty} \de^{t-1} U_i^t $ where $U_i^t$ is player $i$'s payoff in period $t$.

The payoff functions and all parameters, including probabilities in the war lottery, are common knowledge for all players, and actions are immediately observed by all players at the time they occur. We turn next to describing the payoffs.

@@

All the state variables except for the secessionists' status quo payoffs remain unchanged from period to period unless players $p$ or $c$ make an investment.\footnote{We will see below that restrictions on the initial values of the state variables are essential for characterizing our main result, but in general we leave them unrestricted.} To be precise, we have\footnote{Here we anticipate a bit the incentives of the patron and international community to sign the impact of contributions in each case.}
	\begin{equation}
	\begin{aligned}
		L_s^t = L_s^{t-1} - R_p^t + R_c^t \hskip.5in L_g^t = L_g^{t-1} + R_p^t - R_c^t \\
		W_s^t = W_s^{t-1} + R_p^t - R_c^t \hskip.5in W_g^t = W_g^{t-1} - R_p^t + R_c^t \\
		Q_g^t = Q_g^{t-1} - R_p^t + R_c^t \hskip.5in Q_s^t = Q_s^{t-1} - \mu + R_p^t - R_c^t
		\label{eqn:motion}
	\end{aligned}
	\end{equation}

The status quo payoffs for the secessionists are special in that the transition of this state variable from one period to the next involves an automatic reduction by $\mu$. The steady reduction by $\mu$ represents the costs of non-recognition discussed in the introduction. As the economy in the secessionist region deteriorates, so does the standard of living for the secessionists.

@@

If either $s$ or $g$ plays Cede while the other plays Fight or Status Quo, the game enters an absorbing state (i.e. the game ends), with payoffs in every subsequent period given by the corresponding entry in the stage game (Figure 2). We interpret one player ceding as that player ceding the issue of status (independence vs. reunification) in exchange for some set of (relatively small) payments from the opposing player. For example, if the secessionists cede, the secessionist region is reunified with the home state and the payoffs are $L_s$ for the secessionists and $W_g$ for the home state. Therefore, if one player agrees to cede while the other player chooses to remain in the status quo or fight, the result is a negotiated settlement benefiting the player who did not cede.

There are three ways to end up in war: either of the parties may attack first, or both may attack simultaneously. We denote the payoffs of war as a lottery $\Omega$.\footnote{Because the unrecognized state already controls territory and the \emph{de facto} borders are armed, there is likely only a small advantage to be gained by attacking first for either side. Therefore, we argue it is not essential to differentiate between these war scenarios analytically.} This lottery determines whether the secessionists or government wins the war. For simplicity, we assume the outcome of the war is an absorbing state.\footnote{The model can easily be extended to encompass the possibility of an indecisive war.} Outright victory would, among other things, allow an unrecognized state to force recognition by the home state government. Therefore, outright victory requires more than simply securing control of the disputed territory (which the secessionists have already done when the game begins) and requires the ability to impose the terms of settlement. In practice, this likely involves the capture of the home state capital and/or the overthrow of the home state government.

For probabilities $p$ of outright victory and $1-p$ of loss, player $i \in \{s, g\}$ in period $t$ with a fixed cost of war $\zeta_i$ faces war lottery $\omega_i^t \equiv (p (W_i^t-\zeta_i), \left(1-p\right)( L_i^t-\zeta_i))$. $\Omega^t \equiv (\omega_g^t, \omega_s^t).$\footnote{Baseline costs of war are fixed ($\zeta_i$) but additional costs of war based on war's result are captured in $W_i$ and $L_i$. Note that the structure of the war lottery along with the assumption that the outside actors influence the payoffs only and not the probabilities of winning imply that any way in which the patron influences the payoffs from war can be counteracted by the international community through its own investments.} Players are assumed to approach war lotteries as expected values.

@@

We use two binary variables to express the preferences among the three outcomes (status quo, reunification, recognition by the home state) for both $p$ and $c$. $X$ is a binary variable representing reunification: $X=0$ in the status quo and $X=1$ if the secessionists rejoin the home state. Y represents recognition by the home state of the secessionists' independence: $Y=1$ if the home state recognizes the secessionists as independent, $Y=0$ otherwise.

Player $p$'s payoffs in period $t$ are $U_p^t= -\alpha X+\lambda Y -R_p^t$, while the payoffs for player $c$ are $U_c^t= \beta X-\nu Y -R_c^t$, with both payoffs denoted in currency units.\footnote{Note that, in equilibrium, after an absorbing state has been reached, neither outside player will make any investments. Thus the relevant per period payoff will be repeated with $R_p^i = R_c^i = 0$ for $i > t$ if an absorbing state is reached at time $t$.} Taking $\alpha, \ \beta, \lambda,$ and $\nu$ to be positive, these payoffs represent the idea that the patron opposes reunification while player $c$ prefers it, and that the patron prefers recognized independence while player $c$ is averse to the creation of new states.

%The relative strengths of the patron and player $c$'s preferences determine their willingness to spend resources. As we will show, unrecognized statehood emerges as an unhappy but stable middle ground in which all players avoid their least preferred outcomes.

%While in many cases both the central government and the de facto government of the unrecognized states may be relatively autocratic, leaders still rely on some level of public support to remain in power.  If the leaders' actions deviate too far from the public's preferred course, they may be removed from office or forced to bear the costs of additional repression of regime opponents.  Therefore, both the central government and the secessionist elite have incentives to take the public's opinions into account, and the preferences of the public are thus internalized in the payoffs expressed in the model for $c$ and $g$.

\subsubsection{The Low Payoffs from Ceding}
Aside from the central assumption that the government of the home state most prefers reunification and the secessionists most prefer independence accompanied by international recognition, we will for the most part leave the model general enough to incorporate a range of preferences and capabilities for players $g$ and $s$. %Our results will specify parameter ranges for these preferences over which outcomes of interest---such as perpetual unrecognized statehood---can be supported in equilibrium.
We do, however, assert that the payoffs for the party that cedes the issue of status (independence vs. reunification) are consistently low. This reflects a combination of two factors. First, the issue of status is indivisible\footnote{Either the unrecognized state has sovereignty over its territory and is co-equal with the home state, or the secessionist region (and its government) are subordinate to the central government.} and highly valued by each side, making its surrender undesirable. Various forms of ethno-nationalism often motivate secession, and the values attached to independence by secessionists (and to reunification by citizens of the home state) are generally large relative to the values placed on economic prosperity and other goals. Second, many of the payments that could be offered are not credible (e.g. Schultz, 2010).

The difficulty of making credible payments in exchange for status is one clearly demonstrated in the civil war literature (e.g. Licklider, 1995; Walter 1997, 2002; Doyle and Sambanis, 2006; Fearon and Laitin, 2007). Unrecognized states generally constitute ``sons of the soil" conflicts in which the central government cannot credibly commit to preserving the local demographic and political dominance of the secessionist elite once the disputed territory reverts to central government control (Weimer, 1978; Fearon, 2004). While the central government might initially grant the secessionist elite a high level of autonomy in exchange for agreeing to reunification, the level of autonomy is likely to decrease over time, perhaps quite quickly. The payments that can be offered by the unrecognized state to the central government are similarly small or unenforceable: once recognition is achieved, the secessionists have little reason to uphold any prior commitments.

%At the time of secession, ethnic Akbhaz made up a minority of the population of Abkhazia, but by the late 1990s they controlled almost all political posts in the \textit{de facto} government of the region (Cornell 2001; Wooleh 2006). In 2004, the basket of payments offered by the Georgians in exchange for reunification included a provision guaranteeing that ethnic Abkhaz would retain a majority in the regional parliament, even if the return of internally displaced persons (IDPs) once again placed ethnic Abkhaz in a minority demographic position in the region. The promise, however, was not very meaningful. First, even if the promise were upheld, it would still mean a step back from the total dominance the ethnic Abkhaz currently enjoy in the region. Second, if Georgian IDPs returned, they may demand and receive a more equitable system of representation. These concerns are not abstract; this type of reneging has already occurred in cases that did reach settlement.

The case of Gagauzia illustrates the commitment problem nicely. Gagauzia achieved \emph{de facto} independence at the time of the Soviet Union's collapse, but agreed to rejoin Moldova in 1994 as an autonomous region. While Gagauzia was granted substantial autonomy under the Moldovan Law on the Special Legal Status of Gagauzia, when the governor of Gagauzia, Dmitrii Croiter, moved to assert these powers in 1999, the Moldovan government balked. By 2002, Croiter was forced to resign, effectively deposed by the Moldovan government. The Moldovan government jailed a number of other Gagauz politicians, and while \emph{de jure} Gagauz autonomy was enshrined in the Moldovan constitution in 2003, the \emph{de facto} level of autonomy has been limited by continued central government meddling in less-than-free regional elections. The payoffs to Gagauzia for ceding have turned out to be quite low, and a similar fate can rationally be expected by other unrecognized states who choose to cede.
%WE CAN CUT THE GAGAUZ EXAMPLE IF NECESSARY -- BG July 2014

%The payments that can be offered by the unrecognized state to the central government are similarly small or unenforceable. To return to the Abkhaz example: Under a scenario in which Georgia recognizes Abkhazia as an independent state, the ethnically Georgian region of Gali, currently under Abkhaz control, would likely rejoin Georgia, Russian troops would be (at least temporarily) expelled, and compensation might be paid to displaced Georgians, but other side payments are difficult to picture. Abkhazia might promise Georgia privileged access to Abkhaz ports, or promise to keep Russian troops out of its territory permanently, but once recognition is granted, any such promises could easily be reneged upon.

%Furthermore, while relinquishing territorial claims relieves the home state of a persistent source of instability, conceivably supplementing $g$'s payoffs from ceding, it also would increase the probability of future attempts at secession by other regions (Walter 2006). The lack of enforceable side payments and the ambiguous impact on stability make ceding the issue of status a low-payoff option for $g$, just as it is for $s$.

%Among the means through which $p$ or $c$ can spend resources to alter the payoffs to $g$ and $s$ is to enforce bargains and guarantee future concessions by either party, raising the payoffs from ceding. For example, if both the home state and the secessionists prefer reunification with autonomy rights to both war and continued stalemate, $c$ can spend resources to enforce an agreement in which the secessionists are promised specific autonomy rights and $c$ agrees to ensure that these rights are later retained. Should the home state later attempt to revoke the promised autonomy, $c$ can levee sanctions or employ other coercive measures to deter this action. The promises and pitfalls of this approach are discussed in the section on policy implications of the model.

>>>>

The game begins at a status quo in which the secessionist elite controls at least some of the disputed territory but cannot gain international recognition unless the central government cedes its claim to the territory. This condition is archetypical of cases in which a militarily successful war of secession ends in a ceasefire.

@@

\item $c$ chooses an investment level $R_c^t\in \mathbb{R}^+$ to influence the payoffs of $s$ and $g$.

@@

The game continues into the next period if the `inside players'---the secessionists and government---selected (status quo, status quo) or (cede, cede). Otherwise, the game ends in an absorbing state as described in Section~\ref{sec:stage}.

Future payoffs are discounted with a common parameter $\de \in [0,1]$. Therefore payoffs for the entire game for player $i\in \{s, g, p, c\} $ can be expressed by the discounted stream of payments $\Sigma_{t=1}^{\infty} \de^{t-1} U_i^t $ where $U_i^t$ is player $i$'s payoff in period $t$.

The payoff functions and all parameters, including probabilities in the war lottery, are common knowledge for all players.

@@

All the state variables except for the secessionists' status quo payoffs remain unchanged from period to period unless players $p$ or $c$ make an investment. To be precise, we have\footnote{Here we anticipate a bit the incentives of the patron and international community to sign the impact of contributions in each case.}
	\begin{equation}
	\begin{aligned}
		L_s^t = L_s^{t-1} - R_p^t + R_c^t \hskip.5in L_g^t = L_g^{t-1} + R_p^t - R_c^t \\
		W_s^t = W_s^{t-1} + R_p^t - R_c^t \hskip.5in W_g^t = W_g^{t-1} - R_p^t + R_c^t \\
		Q_g^t = Q_g^{t-1} - R_p^t + R_c^t \hskip.5in Q_s^t = Q_s^{t-1} - \mu + R_p^t - R_c^t
		\label{eqn:motion}
	\end{aligned}
	\end{equation}

The status quo payoffs for the secessionists are special in that the transition of this state variable from one period to the next involves an automatic reduction by $\mu$. The steady reduction by $\mu$ represents the costs of non-recognition discussed in the introduction.

@@

If either $s$ or $g$ plays Cede while the other plays Fight or Status Quo, the game enters an absorbing state (i.e. the game ends), with payoffs in every subsequent period given by the corresponding entry in the stage game (Figure 2). We interpret one player ceding as that player ceding the issue of status (independence vs. reunification) in exchange for some set of (relatively small) payments from the opposing player. Therefore, if one player agrees to cede while the other player chooses to remain in the status quo or fight, the result is a negotiated settlement benefiting the player who did not cede.

There are three ways to end up in war: either of the parties may attack first, or both may attack simultaneously. We denote the payoffs of war as a lottery $\Omega$. This lottery determines whether the secessionists or government wins the war. For simplicity, we assume the outcome of the war is an absorbing state. Outright victory would, among other things, allow an unrecognized state to force recognition by the home state government.

For probabilities $p$ of outright victory and $1-p$ of loss, player $i \in \{s, g\}$ in period $t$ with a fixed cost of war $\zeta_i$ faces war lottery $\omega_i^t \equiv (p (W_i^t-\zeta_i), \left(1-p\right)( L_i^t-\zeta_i))$. $\Omega^t \equiv (\omega_g^t, \omega_s^t).$ Players are assumed to approach war lotteries as expected values.

We do not restrict the preferences and capabilities for players $g$ and $s$, with one exception. We assert that the payoffs for the party that cedes the issue of status (independence vs. reunification) are consistently low. This reflects the fact that the issue of status is indivisible and highly valued by each side and that many of the payments that could be offered are not credible (Licklider, 1995; Walter 1997, 2002; Doyle and Sambanis, 2006; Fearon and Laitin, 2007; Schultz, 2010).

@@

We use two binary variables to express the preferences among the three outcomes (status quo, reunification, recognition by the home state) for both $p$ and $c$. $X$ represents reunification: $X=0$ in the status quo and $X=1$ if the secessionists rejoin the home state. Y represents recognition: $Y=1$ if the home state recognizes the secessionists as independent, $Y=0$ otherwise.

Player $p$'s payoffs in period $t$ are $U_p^t= -\alpha X+\lambda Y -R_p^t$, while the payoffs for player $c$ are $U_c^t= \beta X-\nu Y -R_c^t$, with both payoffs denoted in currency units. Taking $\alpha, \ \beta, \lambda,$ and $\nu$ to be positive, these payoffs represent the idea that the patron opposes reunification while player $c$ prefers it, and that the patron prefers recognized independence while player $c$ is averse to the creation of new states.
